---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# Newplot

[![R-CMD-check](https://github.com/Anakamura14/Newplot-Final-Project/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/Anakamura14/Newplot-Final-Project/actions/workflows/R-CMD-check.yaml)

## Rendered Vignette (HTML)

Click below to view the fully rendered vignette online:

**https://htmlpreview.github.io/?https://github.com/Anakamura14/Newplot-Final-Project/blob/main/inst/doc/newplot_vignette.html**

The vignette provides full documentation, usage examples, development notes, and future work.

---

## Overview

**newplot** is an R package designed to make data visualization faster, more intuitive, and more accessible. While **ggplot2** is powerful and flexible, new and intermediate users often struggle with repeated setup code, theme choices, palette selection, and handling grouping variables. Even experienced analysts find themselves writing the same boilerplate code across projects.

**newplot** addresses these challenges by offering:

- `new_plot()`: a simplified wrapper around ggplot2 that automates common aesthetic and structural decisions  
- `new_plot_gadget()`: an interactive Shiny gadget for building visualizations through a graphical interface  
- A consistent, colorblind-friendly default palette  
- Automatic grouping behavior and palette scaling  
- A reproducible, quoted-variable syntax that works cleanly inside scripts and R CMD check environments  

The goal is to help users focus on **interpreting their data**, not wrestling with syntax.

---

## Motivation

The package was created to reduce friction for learners and analysts who want clean, publication-ready graphics without extensive setup. Many users struggle with:

- remembering which aesthetics belong to which geoms  
- choosing appropriate color scales  
- writing repetitive theme code  
- debugging tidy-evaluation errors  
- switching between interactive and script-based workflows  

By automating these steps, **newplot** provides a stable and predictable foundation for visualization.  
The Shiny gadget extends the package for teaching, prototyping, and hands-on exploration by allowing users to visually construct a plot and automatically generate reproducible code.

---

## Installation

Install the development version from GitHub:

```{r}
devtools::install_github("Anakamura14/Newplot-Final-Project")
library(newplot)
```

---

## Basic Usage

```{r}
# Simple scatterplot
new_plot(mtcars, x = "wt", y = "mpg", type = "point")
```

```{r}
# Grouped scatterplot
new_plot(
  iris,
  x = "Sepal.Length",
  y = "Petal.Length",
  group = "Species",
  type = "point",
  palette = "orange"
)
```

```{r}
# Pipe-friendly workflow
mtcars |>
  dplyr::mutate(cyl = factor(cyl)) |>
  new_plot(x = "cyl", y = "mpg", group = "cyl", type = "violin")
```

These examples highlight newplot’s core purpose:
reducing repetitive code while producing clear, consistent, publication-ready figures.

---

## Interactive Shiny Gadget

The `new_plot_gadget()` function creates an interactive visualization builder:

```{r}
new_plot_gadget(iris)
```

The gadget allows users to:

- Select variables, plot types, palettes, and themes
- Preview the plot instantly
- Export the exact `new_plot()` code for reproducibility

This makes the gadget valuable for teaching, demonstrations, and rapid prototyping.

---

## Key Features

The `new_plot()` package streamlines the standard ggplot2 workflow by automatically managing themes, colors, grouping structures, and palette scaling. When no palette is specified, the function defaults to a colorblind-friendly viridisLite scheme, ensuring clear and accessible visualizations. Its tidy-evaluation–compatible design, which uses quoted variable names, supports smooth behavior in scripts, pipelines, and R CMD check environments. The package also includes an interactive Shiny gadget that allows users to explore visualization options visually and export reproducible code. Currently, `new_plot()` supports several commonly used plot types, including point, line, boxplot, and violin, covering a wide range of exploratory and instructional use cases.

---

## Future Development

There are several promising directions for extending the `new_plot()` package. One goal is to expand the range of supported geoms by incorporating histograms, density plots, bar charts, smoothing layers, and other visualization types commonly used in exploratory analysis. Future versions may also introduce more advanced thematic controls, enabling users to apply reusable style templates or organization-specific formatting standards. Additional enhancements to the palette system—such as improved contrast checking, expanded colorblind-friendly options, and dynamic palette scaling—would further strengthen accessibility. The Shiny gadget has significant potential for growth as well, including features such as faceting controls, image export options, and built-in tutorials for instructional settings. Finally, continued refinement of internal validation and automated testing would help prepare the package for broader use, including potential submission to CRAN.

